# GO and KEGG enrichment
```{r echo=F, warning=F}
age_spe_up_gk <- lapply(age_spe_up, function(x) myGK(rownames(x)))
# age_spe_dn_gk <- lapply(age_spe_dn, function(x) myGK(rownames(x)))
# 
# # paste(age_spe_up_gk[["2m"]]$GO$MF$Term, collapse = ", ")
# sapply(age_spe_up_gk, function(x) paste(x$KEGG$Term, collapse = ", "))
# sapply(age_spe_dn_gk, function(x) paste(x$KEGG$Term, collapse = ", "))

```

## Mixed Age and App effects
```{r echo=F, warning=F}
# mixed_eff <- lapply(1:ncol(combn), function(x1) {
#   y1 = sapply(glm.fit, function(x2) {
#     y2 = contrast(x2, list(age = combn[2, x1], group = "APP"), list (age = combn[1, x1], group = "APP")) 
#     c(y2$Contrast, y2$Pvalue)})
#   rownames(y1) = c("contrast", "pval"); t(y1)
# }); names(mixed_eff) <- apply(combn, 2, function(x) paste0(x[2], x[1])) 
```

##### In each gene's GLM return, the "Estimate" column represents each predictor's effect size; the "Pr(>|t|)" column represents the corresponding significance. We constructed a matrix of each gene's estimates on each predictor, and another matrix of the corresponding significance values.
```{r}
fit.qr[["App"]]$coefficients
fit.est <- lapply(fit.qr, function (x) x$coefficients[, "Estimate"])
fit.est <- do.call(rbind, fit.est)
tail(fit.est)
fit.pval <- lapply(fit.qr, function (x) x$coefficients[, "Pr(>|t|)"])
fit.pval <- do.call(rbind, fit.pval)
tail(fit.pval)
```

##### To identify each gene's critical predictors, we binarized the estimate and significance matrices by hard cutoffs. Less stringent cutoffs were applied in a separate analysis. 
```{r}
logit <- apply(fit.pval, 2, function (x) x < 0.05) & apply(fit.est, 2, function (x) abs(x) > 0.2)
tail(logit)
```

##### The APP- and age-related genes were identified as those who passed the cutoff in at least one of the APP- and age-related predictors, respectively. 
```{r}
geneId.age <- geneId[as.logical(rowSums(logit[, grep("age", colnames(fit.pval))]))]
geneId.age
geneId.app <- geneId[as.logical(rowSums(logit[, grep("APP", colnames(fit.pval))]))]
geneId.app
```

##### GO and KEGG enrichment analysis was applied on the APP- and age-related genes
```{r}
gk.app <- myGK(geneId.app)
head(gk.app$BP[, c("Term", "Pvalue")])
head(gk.app$MF[, c("Term", "Pvalue")])
head(gk.app$CC[, c("Term", "Pvalue")])
head(gk.app$KEGG[, c("Term", "Pvalue")])

gk.age <- myGK(geneId.age)
head(gk.age$BP[, c("Term", "Pvalue")])
head(gk.age$MF[, c("Term", "Pvalue")])
head(gk.age$CC[, c("Term", "Pvalue")])
head(gk.age$KEGG[, c("Term", "Pvalue")])
```

##### This gives the overlap between the age- and app-genes.
```{r}
grid.newpage()
draw.pairwise.venn(length(geneId.age), length(geneId.app), length(intersect(geneId.age, geneId.app)), 
                   category = c("Age", "APP"), fill = c("light blue", "pink"))
```
                   
##### To classify the APP-related genes in more detail, we collapsed the binarized matrix gene by gene and grouped genes from the unique collapsed patterns. We will repeat the same procedure for the age-related genes.
```{r}
profile <- logit[geneId.app, grep("APP", colnames(logit))]
# profile <- logit[geneId.age, grep("age", colnames(logit))]
tail(profile)

profile.str <- apply(profile, 1, function (x) paste(x, collapse = "-"))
tail(profile.str)
profile.table <- sort(table(profile.str))
profile.Id <- names(profile.table)
profile.Id
```

##### Genes in each group
```{r}
geneId.profile <- list()
for (idx in profile.Id)
  geneId.profile[[idx]] <- geneId.app[profile.str == idx]
# geneId.profile[[idx]] <- geneId.age[profile.str == idx]
geneId.profile
```

##### A tile plot on the groups with each group's gene numbers
```{r}
tile.dt <- NULL
for (i in 1:length(profile.Id))
  tile.dt <- rbind(tile.dt, as.logical(unlist(strsplit(profile.Id[i], "-"))))
tile.dt <- data.frame(value = c(tile.dt), 
  profile = factor(rep(profile.Id, ncol(tile.dt)), levels = profile.Id),
  group = factor(rep(colnames(profile), each = nrow(tile.dt)), levels = colnames(profile)))

ggplot(tile.dt, aes(x = group, y = profile, fill = value)) + geom_tile(colour = "white") +
  theme_bw() + xlab("") + ylab("") + coord_flip() +
  scale_x_discrete(labels = c("2m:APP", "4m:APP", "5m:APP", "6m:APP")) +
# scale_x_discrete(labels = c("4m:WT", "5m:WT", "6m:WT", "4m:APP", "5m:APP", "6m:APP")) +
  scale_y_discrete(labels = profile.table) +
  scale_fill_manual(values = c("grey80", "firebrick1")) 
``` 

##### We repeated the same procedure to classify age-related genes in different groups.
```{r}
# profile <- logit[geneId.app, grep("APP", colnames(logit))]
profile <- logit[geneId.age, grep("age", colnames(logit))]
tail(profile)

profile.str <- apply(profile, 1, function (x) paste(x, collapse = "-"))
tail(profile.str)
profile.table <- sort(table(profile.str))
profile.Id <- names(profile.table)
profile.Id
```

##### Genes in each group
```{r}
geneId.profile <- list()
for (idx in profile.Id)
# geneId.profile[[idx]] <- geneId.app[profile.str == idx]
  geneId.profile[[idx]] <- geneId.age[profile.str == idx]
geneId.profile
```

##### A tile plot on each group's gene numbers
```{r}
tile.dt <- NULL
for (i in 1:length(profile.Id))
  tile.dt <- rbind(tile.dt, as.logical(unlist(strsplit(profile.Id[i], "-"))))
tile.dt <- data.frame(value = c(tile.dt), 
  profile = factor(rep(profile.Id, ncol(tile.dt)), levels = profile.Id),
  group = factor(rep(colnames(profile), each = nrow(tile.dt)), levels = colnames(profile)))

ggplot(tile.dt, aes(x = group, y = profile, fill = value)) + geom_tile(colour = "white") +
  theme_bw() + xlab("") + ylab("") + coord_flip() +
# scale_x_discrete(labels = c("2m:APP", "4m:APP", "5m:APP", "6m:APP")) +
  scale_x_discrete(labels = c("4m:WT", "5m:WT", "6m:WT", "4m:APP", "5m:APP", "6m:APP")) +
  scale_y_discrete(labels = profile.table) +
  scale_fill_manual(values = c("grey80", "firebrick1")) 
``` 
