## Stratify the adult APP samples
Xulong Wang (xulong.wang@jax.org)

***

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ape)
library(amap)
library(ggdendro)
```

```{r, message=F, warning=F}
rm(list = ls())
setwd("~/Dropbox/GitHub/Ad")
load("data/brain_bc2014.rdt")
load("data/glm_brain.rdt")
source("function.R")
```

##### The batch-corrected gene expression values for the App- and Age-related genes from GLM were used for this analysis.
```{r}
geneId.app = glm$less$app$symbol
geneId.age = glm$less$age$symbol
dt.hc <- dt.bc[unique(c(geneId.app, geneId.age)), ]
dim(dt.hc)
```

##### Two-way hierarchical clustering (HC) was applied on the dataset.
```{r}
hc1 <- hcluster(t(dt.hc), method = "pearson", link = "average")
hc2 <- hcluster(dt.hc, method = "pearson", link = "average")

mycol <- rep("grey50", ncol(dt.hc))
mycol[grep("APP", colnames(dt.hc))] <- "firebrick1"
plot(as.phylo(hc1), edge.width = 2, font = 2, cex = 0.7, label.offset = 1e-4, tip.color = mycol, direction = "downward")
```

##### From the HC dendrogram, we found two distinct clusters of 5 and 6 month APP samples. Why the two adult APP cohorts so different? 
```{r}
group1<- c("APP5m1558.2014", "APP5m2751.2014", "APP5m1633.2014","APP5m1636.2014","APP6m1684.2014","APP5m1648.2014", "APP5m1623.2014")
group2<- c("mouse_3346_6m_APP", "mouse_3440_6m_APP", "mouse_3351_6m_APP", "mouse_6585_6m_APP", "APP5m1647.2014", "APP5m1738.2014")

mycol[colnames(dt.hc) %in% c(group1, group2)] <- "blue"
plot(as.phylo(hc1), edge.width = 2, font = 2, cex = 0.7, label.offset = 1e-4, tip.color = mycol, direction = "downward")
```

##### To identifiy the underlying genes that drove the two clusters, we performed a pairwise t-test for the two groups on the complete geneset.
```{r}
dt.g1 <- dt.bc[, group1]
dt.g2 <- dt.bc[, group2]
colnames(dt.g1) <- paste("group1", colnames(dt.g1), sep = "_")
colnames(dt.g2) <- paste("group2", colnames(dt.g2), sep = "_")
dt.g12 <- cbind(dt.g1, dt.g2)
dim(dt.g12)

fc <- rowMeans(dt.g1) - rowMeans(dt.g2) 
treat <- gsub("^.*(group1|group2).*", "\\1", colnames(dt.g12))
tt.pval <- apply(dt.g12, 1, function(x) pairwise.t.test(x, treat, p.adj = "none")$p.value)
tt.qval <- p.adjust(tt.pval, method = "fdr")
```

##### Here are the genes who fit two criteria: 1. FDR < 0.05; 2. Fold change > 0.2
```{r}
idx <- which(tt.qval < 0.05 & abs(fc) > 0.2)
geneId <- rownames(dt.g12)[idx]
head(geneId)
```

##### GO and KEGG enrichment
```{r}
gk.5m <- myGK(geneId)
head(gk.5m$BP[, c("Term", "Pvalue")])
head(gk.5m$MF[, c("Term", "Pvalue")])
head(gk.5m$CC[, c("Term", "Pvalue")])
head(gk.5m$KEGG[, c("Term", "Pvalue")])
```
