## Identify co-expression gene modules by the App and Age phenotypes
Xulong Wang (xulong.wang@jax.org)

***

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ape)
library(amap)
library(ggplot2)
library(ggvis)
library(dynamicTreeCut)
```

```{r, message=F, warning=F}
rm(list = ls())
setwd("~/Dropbox/GitHub/Ad")
load("./data/brain_bc2014.rdt")
source("function.R")
```

##### Batch-corrected gene expression dataset from GLM was filtered to keep the most variable genes across conditions for downstream analysis. We applied ANOVA first to remove genes that were highly variable among replciates, followed by an explicit selection of highly variable genes among conditons.
```{r}
dt <- dt.bc
aov.pval <- NULL  
for (i in 1:nrow(dt)) {
# if (i %% 1e3 == 0) cat(i, "\n")
  dat1 <- data.frame(y = as.matrix(dt)[i, ], age, group)
  aov <- aov(y ~ age * group, data = dat1)
  aov.pval <- c(aov.pval, min(summary(aov)[[1]][["Pr(>F)"]], na.rm = T))
}
dt.cond <- NULL
dt <- dt[aov.pval < 0.05, ]
for (idx in conditions) dt.cond <- cbind(dt.cond, rowMeans(dt[, uid == idx]))
colnames(dt.cond) <- conditions
dt.cond <- dt.cond[apply(dt.cond, 1, sd) > 0.1, ]
dt <- as.data.frame(dt.cond)
n.gene <- nrow(dt)
geneId <- rownames(dt)
str(dt)
```

##### A similarity measure, which describes the distances between network nodes, is the core for network construction. We constructed a topology overlap matrix (TOM) as the similarity measure based on the assumption that gene expression networks follow the scale free topology rule. The TOM took acount not only distance between two nodes, but also how many nodes they share. Code block below constructs the TOM. 
```{r}
sft <- NULL  # --- Scale free topology
similarity <- cor(t(dt), method = "pearson")
diag(similarity) <- 0
for (beta in seq(1, 31, 2)) {
  adjacency <- abs(similarity)^beta
  k1 <- rowSums(adjacency)
  discretized.k = cut(k1, 20)  # breaks:20
  dk = tapply(k1, discretized.k, mean)
  p.dk = tapply(k1, discretized.k, length) / length(k1)
  dk = dk[! is.na(dk)]
  p.dk = p.dk[! is.na(p.dk)]
  fit <- summary(lm(log10(p.dk) ~ log10(dk)))
  sr2 <- -sign(fit$coefficients["log10(dk)", "Estimate"]) * fit$r.squared
  sft <- rbind(sft, c(beta, sr2))
}
plot(sft, main = "Scale free topology", xlab = "beta", ylab = "signed R2")
sft.beta <- 27
adjacency <- abs(similarity)^sft.beta
tom <- matrix(0, n.gene, n.gene)
rownames(tom) <- colnames(tom) <- geneId
k1 <- rowSums(adjacency)
for (i in 2:n.gene) {
# if (i %% 1e2 == 0) cat(i, "\n")
  for (j in 1:(i-1)) {
    kij <- min(k1[c(i, j)])
    lij <- sum(adjacency[i, ] * adjacency[j, ])
    wij <- (lij + adjacency[i, j]) / (kij + 1 - adjacency[i, j])
    tom[i, j] <- wij
  }
}
diag(tom) <- 1
tom[upper.tri(tom)] <- t(tom)[upper.tri(tom)]
diss.tom <- 1 - tom
```

##### To detect co-expression gene modules, hierarchical clustering was applied on the TOM. Modules were identified by cutting the hierarchical clustering dendrogram into branches.
```{r}
dendro <- hclust(as.dist(diss.tom), method = "average")
branch <- cutreeDynamic(dendro = dendro, distM = diss.tom, minClusterSize = 30)
names(branch) <- geneId
table(branch)
```

##### Genes in each modules were retrieved. 
```{r}
geneId.network <- list()
for (idx in 1:8) geneId.network[[idx]] <- names(branch[branch == idx])
geneId.network[[6]]
```

##### Expression patterns of genes in each module can be represented by a psedo-gene, which was computed as the first principal component of the module genes's expression profiles. 
```{r}
eigene <- NULL
branchId <- names(table(branch))[-1]
for (idx in branchId) {
  dt1 <- dt[branch == idx, ]
  dt1 <- t(apply(dt1, 1, scale))
  svd1 <- svd(dt1)
  eigene <- rbind(eigene, svd1$v[, 1])
}
dimnames(eigene) <- list(branchId, conditions)
eigene
```

##### What the psedo-genes look like?
```{r, warning=F}
module <- paste("Module", branchId, sep = "")
group <- factor(conditions, levels = conditions)
gdt <- data.frame(value = c(eigene), module = rep(module, 8), group = rep(group, each = 8))
gdt$geno <- factor(gsub("^.*(WT|APP).*", "\\1", gdt$group), levels = c("WT", "APP"))
ggplot(gdt, aes(x = group, y = value, fill = geno)) + 
  geom_bar(stat = "identity") + facet_grid(. ~ module) +
  scale_fill_manual(values = c("red", "blue")) +
  theme_bw() + xlab("") + ylab("") + 
  theme(axis.text.x = element_text(angle = 90))
```

##### Are we getting the reasonable gene modules? This shows expression profiles of genes in module 6. 
```{r}
x.dt = as.matrix(dt[branch == 6, ])
x.dt <- t(apply(x.dt, 1, scale))
gdt <- data.frame(value = c(x.dt), gene = rep(rownames(x.dt), 8), group = rep(group, each = nrow(x.dt)))
gdt %>% ggvis(~group, ~value, stroke = ~gene) %>% layer_lines() %>% hide_legend("stroke") %>%
  add_tooltip(function (x) paste("Gene: ", x$gene), "hover")
```

##### GO and KEGG enrichment
```{r}
geneId <- geneId.network[[6]]
gk.mod6 <- myGK(geneId)
head(gk.mod6$BP[, c("Term", "Pvalue")])
head(gk.mod6$MF[, c("Term", "Pvalue")])
head(gk.mod6$CC[, c("Term", "Pvalue")])
head(gk.mod6$KEGG[, c("Term", "Pvalue")])
```
